### 2019-06-18 2-Chapters/1-Chapter/chapter1.tex
Almost okay for summary/history of Ch4, Ch5, Ch6

% The opposite: decentralized broadcasting, and the age of IoT
Finally, a third kind of systems are of the opposite kind, and can be designed as decentralized broadcasting: an antenna is still in charge of many devices, but such devices can only send up-link packets, and the only down-link data they can receive are short acknowledgements sent from the base-station to indicate success or failure of an up-link packet.
This family of wireless systems are referred to as Internet of Things,
and a typical example of application is for sensor networks.\\
\indent
For the future development of smart grids, smart cities, smart homes, or smart agriculture, sensor networks are promised to be widely deployed.
Two examples of future applications that are already in deployment, in France or other countries, are connected buildings and connected agriculture.
For buildings, the main goal is to reduce the cost of heating empty buildings and use temperature sensor networks to get accurate and regular data about the temperature in each room and each floor, and let the centralized heat control optimize its cost and energy consumption.
For agriculture, one example can be to equipped every cow (in large farms) with a sensors that emits biological information, such as body temperature or stress level etc, in order to optimize the time of milking and monitors the health of the animals.\\
%
% OK commencer par un paragraphe qui explique qu'on étudie les télécom dans le but de les rendre automatiquement plus efficace, plus verte, plus adaptative
\indent
This third kind of wireless systems are characterized by their decentralized nature: due to the really limited information that the base station can send to the devices, it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
In the present and future IoT networks, many devices of heterogenous nature are using the same antenna for many different applications.
A common hypothesis is that such IoT devices have a strong constraint on their power consumption, as most of them will be deployed without a direct power access and will use a battery, which duration should be maximized (typically more than 10 years).
Another common constraint for IoT devices is their low duty cycle, as most applications target one to a few messages sent by day, in strike opposition to the high data-rate pursued for centralized systems such as 4G/5G and WiFi.
%
Many different standards for IoT networks have been proposed,
and they consist in a specification for both the \emph{PHY}sical
and the Medium Access Control (\emph{MAC}) layer.
To quote some examples of standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications, while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications.
We refer to the survey \cite{Centenaro16} for more details, and references in \cite{Azari18} or our recent article \cite{MoyBesson2019}.

@@

\textbf{The spectrum scarcity issue.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, limiting the possibility of adding new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below, only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz}, while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz}, is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication) to space research, inter-satellite, mobile telephony and many other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the \href{https://www.fcc.gov/}{Federal Communications Commission} in the USA (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}, previously known as \emph{``TéléDiffusion de France''}), as well as different independent measurement campaigns, found that most radio frequency spectrum was inefficiently utilized,
meaning that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{valenta2010survey} for a survey on spectrum utilization in Europe.

@@

\paragraph{From Machine Learning to Multi-Armed Bandits.}
%
We can narrow down the first field of study of this thesis, steps by steps.
Starting from the general concept of Machine Learning (ML), we focused our interest to statistical learning, then sequential learning, then Reinforcement Learning (RL), then RL with limited feedback and finally to Multi-Armed Bandits (MAB) learning.
As illustrated in Figure~\ref{fig:1:MainTypesOfMachineLearning}, there are other types of ML, that we do not focus on in this thesis.
The reference book on RL is \cite{SuttonBarto2018}, so let us quote the definition of RL given by R. Sutton and A. Barto:
\emph{``Reinforcement learning is learning what to do -— how to map situations to actions -— so
as to maximize a numerical reward signal. The learner is not told which actions to
take, but instead must discover which actions yield the most reward by trying them''}.
% - apprentissage statistique (machine learning) > séquentiel > par renforcement > information partielle (bandits)

@@

\paragraph{From IT to Cognitive Radio.}
%
Similarly, we can narrow down the second field of study of this thesis, steps by steps:
from Information Technologies (IT), to telecommunications, to wireless communications, then to Software Defined Radio (SDR),
and finally to Cognitive Radio (CR).
%
The transition from the historical approach of hardware-based radio to SDR architectures and is a gradual process that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (e.g. mixers, filters, amplifiers, modulators/demodulators, detectors, etc) are instead implemented by means of software on a personal computer or embedded system, with the exception of a unique pair of receiver/transmiter antennas.
Even though the SDR paradigm was initiated by the USA defense research, over the past years the industry has started to be interested by SDR, and CR have got significant attention from the academia and industry as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
CR is not a standard technology, and as such it does not have a single definition, so let us start by quoting the definition of two researchers whose works were the roots of the development of CR, in the last $20$ years.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}
    \item
    Joseph Mittola in $1999$ said that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned previous works have shown that MAB algorithms can be applied with success for the OSA problem.
But if we consider CR-ready devices that cannot perform sensing, such as low-cost and low energy consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there are no longer a distinction between PU and SU.
%
The specificities of IoT networks can be listed as follows.
\begin{itemize}
    \item
    most IoT networks run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, with low material and computational capacities,
    meaning that they are not equipped with any sensing capabilities.
    However, they are equipped with a transceiver as well as with a receiver,
    and with low-complexity storage and computation capacities,
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low duty cycles (only a few messages every minute) to very low duty cycles (only a few messages a day).
\end{itemize}
\TODOL{FIXME more specificities? Est-ce que je dois mentionner d'autres contraintes ? Il me faudrait aussi une ou deux références genre survey des standards IoT}

@@

\paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many different MAB algorithms, we are also interested by the question of how practitioner can chose the one she will implement in a given IoT object.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well-known existing algorithms in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}, where it appears that simple algorithms such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal} are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is an online algorithm selection, consisting in aggregating a set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.

@@

Verifying the potential gain of performance brought by these MAB algorithms in numerical simulations is a first step, but it is satisfactory to also verify our results on a real-world proof-of-concept (PoC).
In Section~\ref{sec:4:firstModel}, we propose the first PoC of using decentralized MAB learning in a simplified ``IoT'' like wireless network.
We were able to implement a network with a base station which receives incoming messages from different end-devices and replies with acknowledgements, that are used as feedback by the devices and let them learn an efficient spectrum access scheme.
% Verify this first model empirically: proof of concept
This work was again implemented with R. Bonnefoi, and was demonstrated during three days at the IEEE ICT conference in Saint-Malo, France, in June $2018$ \cite{Besson2018ICT}.

We then consider an extension of our first model, following the idea of packet retransmission that is at the root of the ALOHA protocol \cite{Abramson1970,Roberts75}.
This second model proposes to also use decision making at the second layer of the MAC protocol, that is to also use RL and MAB algorithms, in order to learn the best channel to use for retransmitting any packet that suffered from a collision at its first transmission.
We present the model in Section~\ref{sec:4:retransmissions}, along with mathematical justifications of the potential of using this second-stage MAB learning for retransmissions, and numerical illustrations.
This shows that using learning in a decentralized way on the devices' side is again a powerful idea under this model with retransmissions, and we show as well as the simplest of the different heuristics we compared is efficient and allow the devices to learn to behave almost optimally.
This third contribution concluded our collaboration with R. Bonnefoi, and was presented at the MoTION workshop during the IEEE WCNC conference in Marrakech, Morocco, in April $2019$ \cite{Bonnefoi2019WCNC}.

\paragraph{Chapter~\ref{chapter:5}.}
%
In the two models presented above, the core concept is to let all the dynamic devices of a IoT network run a fully decentralized learning algorithm,
each of them targeting their local objective of maximizing their cumulated reward (in a selfish way), in order to optimize the complete system.
It is well known in the game theory literature that playing selfishly can be disastrous for the centralized performance measure: for instance, one can think of the prisoner dilemma.
So it is quite surprising that our numerical simulations as well as the realistic PoC showed that decentralized selfish MAB learning lead to efficient

Selfish
\TODOL{Je dois faire ça d'ici mercredi matin}

Understand Selfish in the simpler framework of multi-player MAB
\TODOL{Je dois faire ça d'ici mercredi matin}

In the OSA setting, we failed
\TODOL{Je dois faire ça d'ici mercredi matin}

But we proposed new algorithms for this problem, and relaunched the research on this model \cite{Besson2018ALT}
\TODOL{Je dois faire ça d'ici mercredi matin}

@@

\paragraph{Chapter~\ref{chapter:6}.}
%
FIXME explain non stationarity !

>>>>

% The opposite: decentralized broadcasting, and the age of IoT
Finally, a third kind of systems are of the opposite kind, and can be designed as decentralized broadcasting: an antenna is still in charge of many devices, but such devices can only send up-link packets, and the only down-link data they can receive are short acknowledgements sent from the base-station to indicate success or failure of an up-link packet.
This family of wireless systems are referred to as Internet of Things,
and a typical example of application is for sensor networks.\\
\indent
For the future development of smart grids, smart cities, smart homes, or smart agriculture, sensor networks are promised to be widely deployed.
Two examples of future applications that are already in deployment, in France or other countries, are connected buildings and connected agriculture.
For buildings, the main goal is to reduce the cost of heating empty buildings and use temperature sensor networks to get accurate and regular data about the temperature in each room and each floor, and let the centralized heat control optimize its cost and energy consumption.
For agriculture, one example can be to equipped every cow (in large farms) with a sensors that emits biological information, such as body temperature or stress level etc, in order to optimize the time of milking and monitors the health of the animals.\\
%
% OK commencer par un paragraphe qui explique qu'on étudie les télécom dans le but de les rendre automatiquement plus efficace, plus verte, plus adaptative
\indent
This third kind of wireless systems are characterized by their decentralized nature: due to the really limited information that the base station can send to the devices, it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
In the present and future IoT networks, many devices of heterogeneous nature are using the same antenna for different applications.
A common hypothesis is that such IoT devices have a strong constraint on their power consumption, as most of them will be deployed without a direct power access and will use a battery, which duration should be maximized (typically more than 10 years).
Another common constraint for IoT devices is their low duty cycle, as most applications target one to a few messages sent by day, in strike opposition to the high data-rate pursued for centralized systems such as 4G/5G and WiFi.
%
Many different standards for IoT networks have been proposed,
and they consist in a specification for both the \emph{PHY}sical
and the Medium Access Control (\emph{MAC}) layer.
To quote some examples of standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications, while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications.
We refer to the survey \cite{Centenaro16} for more details, and references in \cite{Azari18} or our recent article \cite{MoyBesson2019}.

@@

\textbf{The spectrum scarcity issue.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, limiting the possibility of adding new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below, only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz}, while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz}, is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication) to space research, inter-satellite, mobile telephony and a large number of other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the \href{https://www.fcc.gov/}{Federal Communications Commission} in the USA (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}, previously known as \emph{``TéléDiffusion de France''}), as well as different independent measurement campaigns, found that most radio frequency spectrum was inefficiently utilized,
meaning that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{valenta2010survey} for a survey on spectrum utilization in Europe.

@@

\paragraph{From Machine Learning to Multi-Armed Bandits.}
%
We can narrow down the first field of study of this thesis, steps by steps.
Starting from the general concept of Machine Learning (ML), we focused our interest to statistical learning, then sequential learning, then Reinforcement Learning (RL), then RL with limited feedback and finally to Multi-Armed Bandits (MAB) learning.
As illustrated in Figure~\ref{fig:1:MainTypesOfMachineLearning}, there are other types of ML, that we do not focus on in this thesis.
The reference book on RL is \cite{SuttonBarto2018}, so let us quote the definition of RL given by R. Sutton and A. Barto:
\emph{``Reinforcement learning is learning what to do - how to map situations to actions - so
as to maximize a numerical reward signal. The learner is not told which actions to
take, but instead must discover which actions yield the most reward by trying them''}.
% - apprentissage statistique (machine learning) > séquentiel > par renforcement > information partielle (bandits)

@@

\paragraph{From IT to Cognitive Radio.}
%
Similarly, we can narrow down the second field of study of this thesis, steps by steps:
from Information Technologies (IT), to telecommunications, to wireless communications, then to Software Defined Radio (SDR),
and finally to Cognitive Radio (CR).
%
The transition from the historical approach of hardware-based radio to SDR architectures and is a gradual process that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, amplifiers, modulators/demodulators, detectors, etc) are instead implemented by means of software on a personal computer or embedded system, with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the USA defense research, over the past years the industry has started to be interested by SDR, and CR have got significant attention from the academia and industry as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
CR is not a standard technology, and as such it does not have a single definition, so let us start by quoting the definition of two researchers whose works were the roots of the development of CR, in the last $20$ years.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}
    \item
    Joseph Mittola in $1999$ said that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned previous works have shown that MAB algorithms can be applied with success for the OSA problem.
But if we consider CR-ready devices that cannot perform sensing, such as low-cost and low energy consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there are no longer a distinction between PU and SU.
%
The specificities of IoT networks can be listed as follows.
\begin{itemize}\tightlist
    \item
    most IoT networks run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, with low material and computational capacities,
    meaning that they are not equipped with any sensing capabilities.
    However, they are equipped with a transceiver as well as with a receiver,
    and with low-complexity storage and computation capacities,
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low duty cycles (only a few messages every minute) to very low duty cycles (only a few messages a day).
\end{itemize}
\TODOL{FIXME more specificities? Est-ce que je dois mentionner d'autres contraintes ? Il me faudrait aussi une ou deux références genre survey des standards IoT}

@@

\paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many different MAB algorithms, we are also interested by the question of how practitioner can chose the one she will implement in a given IoT object.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well-known existing algorithms in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}, where it appears that simple algorithms such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal} are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is an online algorithm selection, consisting in aggregating a set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of Cognitive Radio, where most previous research works only considered the $\UCB_1$ algorithm without justifying this choice rigorously,
and was presented at the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

Verifying the potential gain of performance brought by these MAB algorithms in numerical simulations is a first step, but it is satisfactory to also verify our results on a real-world proof-of-concept (PoC).
In Section~\ref{sec:4:firstModel}, we propose the first PoC of using decentralized MAB learning in a simplified ``IoT'' like wireless network.
We were able to implement a network with a base station which receives incoming messages from different end-devices and replies with acknowledgements, that are used as feedback by the devices and let them learn an efficient spectrum access scheme.
% Verify this first model empirically: proof of concept
This is again the result of a fruitful collaboration with R. Bonnefoi, and this work was demonstrated during three days at the IEEE ICT conference in Saint-Malo, France, in June $2018$ \cite{Besson2018ICT}, and later at the IEEE WCNC conference in Marrakech, Morocco, in April $2019$ \cite{Besson2019WCNC}.

We then consider an extension of our first model, following the idea of packet retransmission that is at the root of the ALOHA protocol \cite{Abramson1970,Roberts75}.
This second model proposes to also use decision making at the second layer of the MAC protocol, that is to also use RL and MAB algorithms, in order to learn the best channel to use for retransmitting any packet that suffered from a collision at its first transmission.
We present the model in Section~\ref{sec:4:retransmissions}, along with mathematical justifications of the potential of using this second-stage MAB learning for retransmissions, and numerical illustrations.
This shows that using learning in a decentralized way on the devices' side is again a powerful idea under this model with retransmissions, and we show as well as the simplest of the different heuristics we compared is efficient and allow the devices to learn to behave almost optimally.
This third contribution concluded our collaboration with R. Bonnefoi, and was presented at the $1^{\text{st}}$ MoTION workshop, during the IEEE WCNC $2019$ conference \cite{Bonnefoi2019WCNC}.

\paragraph{Chapter~\ref{chapter:5}.}
%
In the two models presented above, the core concept is to let all the dynamic devices of a IoT network run a fully decentralized learning algorithm,
each of them targeting their local objective of maximizing their cumulated reward (in a selfish way), in order to optimize the complete system.
It is well known in the game theory literature that playing selfishly can be disastrous for the centralized performance measure: for instance, one can think of the prisoner dilemma.
So it is quite surprising that our numerical simulations as well as the realistic PoC showed that decentralized selfish MAB learning lead to efficient coordination between devices.

We first tried to analyze the performance of this decentralized algorithm, that we refer to as Selfish, in the model of Section~\ref{sec:4:firstModel},
but due to the random number of active devices at each time step (as soon as the probability $p$ of activation is $p < 1$), we were unable to propose a clean analysis.
%
That is why in Chapter~\ref{chapter:5}, we relax the hypothesis of having $M \gg K$ devices in a network with $K$ orthogonal wireless channel, equivalent to having $p < 1$, and we consider the case of devices that each communicate at every step (\ie, $p=1$), and so only up-to $M \leq K$ devices.
We present the studied model in Section~\ref{sec:5:model}, that comes with different variants depending on the feedback level, and cover both the OSA case (\ie, with sensing information) or the IoT case (\ie, without sensing feedback).
The goal was to understand the heuristic of Chapter~\ref{chapter:4}, Selfish, in the simpler framework of multi-player MAB, which has been studied before \cite{Zhao10,Anandkumar10,Anandkumar11} for the case of OSA.

On the first hand, in the OSA setting, we failed to obtain positive result for the Selfish policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability. This result was later confirmed and generalized by \cite{BoursierPerchet18}.
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving over previous state-of-the-art results.
%
Unfortunately this strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

Our work on multi-player bandit was presented at the ALT conference in Lanzarote, Spain, in April $2019$ \cite{Besson2018ALT},
%
%
and it relaunched the research on this model as a few independent teams started to work on different questions after its publication.
%
Since last year, it was proven that the ``no sensing'' case (\ie, IoT) is essentially not harder that the ``sensing'' (\ie, OSA) case, from a theoretical point-of-view, by E. Boursier and V. Perchet \cite{BoursierPerchet18}.
Even if their approach is costly empirically, the result is surprising and very interesting.
We discuss it, along with other recent research works such as \cite{LugosiMehrabian18}, in Section~\ref{sec:5:literatureReviewOtherModels}.
%
Moreover, the case of different means of arms for the different players was studied in \cite{KaufmannAbbas19}, and this generalization of our model is very interested for CR applications, as the mean of an arm is the average quality (\eg, availability) of a radio channel, and as such it has no reason to be uniform among end-devices (\ie, players).
%
Finally the adversarial case of multi-player bandit has started to be studied, independently by \cite{bande2019adversarial} and \cite{AlaturLevyKrause19}.

@@

As explained above, the model of Chapter~\ref{chapter:4} was found intractable to analyze because we consider an IoT network with many end-devices, all following random activation patterns.
Generalizing to different activation probabilities would lead to a model even harder to analyze, and enforcing at most $M \leq K$ devices is not really realistic for IoT networks, but leads to the model of Chapter~\ref{chapter:5}.
%
On the first hand, if the activation pattern of all devices could be fixed, for instance based on a centralized affectation of the devices to different time slots, then the model of multi-player bandit from Chapter~\ref{chapter:5} can be used to let the devices learn an optimal orthogonal affectation of $M \leq K$ devices in the $K$ channels.
However, it is hard to argue that this hypothesis of a centralized time schedule for the end-devices can be realistic, because we study the case of decentralized learning for OSA and IoT precisely in order to avoid any central control of the devices by the base station, as we explained above.
%
On the other hand, FIXME explain why is it interesting to study piece-wise stationary problems from the point-of-view of one player...

\paragraph{Chapter~\ref{chapter:6}.}
%
% FIXME explain non stationarity !
%
In order to also understand precisely how MAB algorithms behave under non-stationary environment, we studied the literature on adversarial and on non-stationary MAB, for both cases of slowly-varying and abruptly-changing environments.
We begin our last Chapter~\ref{chapter:6} by reviewing the existing works,
and we chose to focus on the piece-wise stationary problem.
Two main families of algorithms have been proposed for this problem,
that usually consist in combining an efficient policy designed for the stationary MAB problem (\eg, \UCB{} or \klUCB) and a way to adapt to changes in the arms distributions.
Passively adaptive policies use a window of a fixed or evolving size, or a discount factor to slowly forget about the past observations,
while actively adaptive policies use a statistic test to detect the change-point (\ie, the time steps between two consecutive stationary intervals).
As the recent literature showed that the later approach is usually more competitive, and usually obtains better results from both empirical and theoretical aspects, we chose to develop our own actively adaptive algorithm.

Following two recent previous works \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, we combine an efficient index policy (\klUCB) and an efficient change-point detection test, under the assumption of bounded rewards.
Building on very recent results on the Generalized Likelihood Ratio test (GLRT) for Gaussian and sub-Gaussian variables \cite{Maillard2018GLR}, we instead focus on bounded rewards and Bernoulli distributions.
Using the fact that bounded variables in $[0,1]$ are not only $1/4$ sub-Gaussian but also sub-Bernoulli, we prove the first finite-time guarantees for the GLRT for bounded variables.
We first show finite-time bounds on both the false alarm probability of our test, and its detection delay, under mild assumption on the lengths of the stationary sequences.
Our algorithm, denoted \GLRklUCB, is then presented in two variants, whether change-points are local (\ie, only one arm mean changes at each change-point) or global (\ie, possibly all the arm means change at a time).
%
We prove that by combining the asymptotically optimal policy \klUCB, designed for stationary problems, and our new analysis of the GLRT for bounded rewards, we obtain state-of-the-art guarantees on the regret of our proposed algorithm.
%
The performance of \GLRklUCB{} is also illustrated on numerical experiments on synthetic data, where it is shown to outperforms all the passively adaptive policies as well as the previous actively adaptive policies.
%
This last chapter is based on our last work, since October $2018$, and it first lead to a publication at the GRETSI conference in Lille, France, in August $2019$ \cite{Besson2019Gretsi}.
Our work also lead to the article \cite{Besson2019GLRT}, that will be partially rewritten and completed with more recent results and submitted soon to a journal (probably the Journal of Machine Learning Research) in fall $2019$.
